Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2022-03-21 analysis/paper/shceof.tex
Render

\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

% Pandoc citation processing

\usepackage{gensymb}
\usepackage{subfig}
\usepackage[inline]{showlabels}
\usepackage{chngcntr}
\usepackage{natbib}
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\title{Revisiting the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation using complex Empirical Orthogonal Functions \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }

@@

\begin{verbatim}
## Spherical geometry (s2) switched off
\end{verbatim}

@@

The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures are highly relevant for surface impacts and could be related to high-impact climate extremes.
Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through latitudinal transport of heat, humidity, and momentum \citep{trenberth1980}.

One of the main features of the zonally asymmetric circulation are the Pacific--South American Patterns \citep{mo2001}, which are associated with precipitation anomalies in South America, particularly in the September-October-November trimester (cita).
Besides, \citet{raphael2007} suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies.
\citet{irving2015} showed that planetary wave amplitude in general is associated sea ice concentration, as well as temperature and precipitation anomalies.

Some of these patterns of variability associated to the zonally asymmetric circulation appear to have experienced secular changes, although to our knowledge this has not been extensively studied.
For instance, \citet{raphael2003} found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.

One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude \citep[e.g.][]{vanloon1972, trenberth1980, turner2017}.
This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (\citet{vanloon1972}).
However, this methodology relies on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
In the case of the wave 3, for example, \citet{trenberth1985} observed that it played a role in blocking events, but this was mostly due to increased amplitude of a longitudinally localised wave train instead of an hemispheric-scale zonal wave 3.

The fourier technique can be generalised to integrate all planetary wave intensity irregardless of wave number by computing the wave envelope as done by \citet{irving2015}.
This makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.

Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation \citep{fogt2020} followed by the two PSA \citep{mo2001}.
The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
These patterns are derived by applying EOF to temporal anomalies, but \citet{raphael2003} applied EOF methods specifically to zonal anomalies.

EOFs are more flexible than Fourier decomposition modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
A single EOF can also represent a mixture of two or more physical modes.
The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
\citet{irving2016} characterises this pattern in a Fourier framework.
Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle.
As a result, they reprojected meridional wind fields so that the path of the PSA laid on the equator.
On this new projection, they could identify the PSA using Fourier decomposition.

Finally, another, hybrid, methodology commonly used consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
The Marshal SAM Index \citep{marshall2003}, for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following \citet{gong1999}.
\citet{raphael2004} used a similar methodology to describe the wave 3 in the Southern Hemisphere.
Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean.
This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field.
Similarly, \citet{hobbs2010} propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
These methods are grounded on other methods to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
The use of data of only specific regions or locations makes them suitable for easy computation and for extending back into the data-poor past, but also makes them not very robust and unable to capture non-stationary patterns.

Our objective is to improve our understanding of planetary waves in the Southern Hemisphere using an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
We use complex Empirical Orthogonal Functions \citep{horel1984} as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.

We apply this method to characterise the Springtime zonally asymmetric circulation of the extratropical Southern Hemisphere.
To take vertical distribution into account, we analyse both the troposphere and the stratosphere.

In Section \ref{methods} we describe the methods.
In Section \ref{spatial}, we analyse the spatial patterns of each complex EOF.
In Section \ref{regressions} we study the spatial regressions with geopotential height and temperatuer.
In Section \ref{other-variables}, we analyise the relationship between cEOF1 and Ozone, and between EOF2 and the Southern Annular Mode and the PSA.
In Section \ref{tropical} we look for tropical forcings that explain the variability of each cEOF.
Finally, in Section \ref{precipitation} we show the relationship between these modes of variabiltiy and precipitation in South America and Oceania.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979-2019.
Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends.
Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK \citep{fishpack} and horizontal wave activity fluxes was computed following \citet{plumb1985}.
We used Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2.5\degree~resolution in latitude and longitude.
This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

@@

The amplitude of the zonal waves was obtained through the Fourier transform of the spatial field at each latitude circle.
We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) each variable between 75°S and 45°S for each SON and extracting the wave-1 component of the Fourier spectrum.
We chose this latitude band because it is wide enough to capture most of the relevant anomalies for southern hemisphere mid-latitudes.

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according to \citet{baldwin2009}.
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by \citet{campitelli2021}.
Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated following \citet{mo2001} as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons.
In our study, only the SON values were used.

Linear trends were computed by Ordinary Least Squares (OLS) and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom \citep{wilks2011r}.

@@

\begin{figure}
\centering
\includegraphics{../figures/eof-naive-1.pdf}
\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).}
\end{figure}

In traditional EOF analysis zonal waves appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase \citep{horel1984}.
For instance, Figure~\ref{fig:eof-naive} shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
A similar EOF structure can be seen in 200 hPa (not shown).
Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, considering phase-varying structures \citep{horel1984}.
This method involves computing EOF on the analytic representation of the original field.
This representation is a complex field in which the real part is the original variable and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
The Hilbert transform is usually understood in terms of time-varying signal.
However, in this work we apply the Hilbert transform to each latitude circle at each moment in time.
Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.

The result of the cEOF methodology is a set of complex spatial patterns and complex time series.
The real and imaginary part of each spatial pattern represent two phases wave-like spatial patterns that are in quadrature.
The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.

@@

\caption{\label{tab:corr-ceof-splitted}Coefficient of determination (\(R^2\)) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separately for each level. p-values lower than 0.01 in bold.}
\centering
\begin{tabular}[t]{l>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{} & \multicolumn{3}{c}{50 hPa} \\
\cmidrule(l{3pt}r{3pt}){2-4}
200 hPa & cEOF1 & cEOF2 & cEOF3\\
\midrule
cEOF1 & \textbf{0.29} & 0.01 & 0.03\\
cEOF2 & 0.00 & \textbf{0.59} & 0.02\\
cEOF3 & 0.00 & 0.00 & 0.01\\
\bottomrule
\end{tabular}
\end{table}

The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S at 50 and 200 hPa, representing a stratsopheric and upper troposhperic level respectively.
Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
There's a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).

Both the spatial pattern similarity and the high temporal correlation of cOEFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability.
This motivates the decision of performing complex EOF jointly between levels.
The computation of the cEOFs was carried out using data from both levels at the same time, therefore, each cEOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

The phase of principal components is defined up to an additive constant.
For real principal components, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
For complex principal components, it can be any number between 0 and \(2\pi\) \citep{horel1984}.
Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
This procedure does not create a spurious correlation, it only takes a known relationship and aligns it with a specific phase.
For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
This also nearly minimises the correlation with the imaginary part.
For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.

@@

We performed linear regressions to quantify the association between the cEOFs and other variables.
For each cEOF, we create regression maps by fitting a multiple linear model involving both the real and the imaginary part.
To obtain the linear coefficients of a variable \(X\) (geopotential, temperature, precipitation, etc\ldots) with the Imaginary and Real parts of each cEOF we fit the equation

@@

where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients for Imaginary and Real parts respectively, \(X_0\) and \(\epsilon\) are the constant and error terms respectively.

We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate \citep{benjamini1995, wilks2016} to avoid misleading results from the high number of regressions \citep{walker1914, katz1991}.

@@

We performed all analysis in this paper using the R programming language \citep{rcoreteam2020}, using data.table \citep{dowle2020} and metR \citep{campitelli2020} packages. All graphics are made using ggplot2 \citep{wickham2009}. We downloaded data from reanalysis using the ecmwfr package \citep{hufkens2020} and indices of the ENSO with the rsoi package \citep{albers2020}. The paper was rendered using knitr and rmarkdown \citep{xie2015, allaire2020}.

@@

\begin{figure*}
\includegraphics{../figures/extended-series-1} \caption{Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019). Dark straight line is the OLS trend. Black horizontal and vertical line mark the mean value and range of each time series respectively.}\label{fig:extended-series}
\end{figure*}

@@

The cEOF1 (Fig.~\ref{fig:ceofs-1} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
The cEOF2 (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
This modulated amplitude is especially apparent at 200 hPa.
There is no apparent phase shift but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.

All cEOFs have non zero mean (Fig. \ref{fig:extended-series}) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
There are no significant simultaneous correlation between the time series.
Both cEOFs show year-to-year variability but show no evidence of decadal variability.

>>>>

\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}

@@

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{gensymb}
\usepackage{subfig}
\usepackage[inline]{showlabels}
\usepackage{chngcntr}
\usepackage{natbib}
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\begin{document}

\title{Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }

@@

@@

The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts (e.g. Hoskins and Hodges 2005).
They strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum (K. E. Trenberth 1980; M. N. Raphael 2007) and could even be related to the occurrence of high-impact climate extremes (Pezza, Rashid, and Simmonds 2012).

Zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential heights or sea-level pressures at each latitude (e.g. van Loon and Jenne 1972; K. E. Trenberth 1980; Turner et al. 2017).
This approach suggests that zonal waves 1 and 3 explain almost 99\% of the total variance in the annual mean 500-hPa pattern at 50ºS (van Loon and Jenne 1972).
K. F. Trenberth and Mo (1985) concluded that wave 3 plays a role in the development of blocking events.
In addition, previous works have identified at extratropical and subpolar latitudes, wave-like patterns with dominant wavenumbers 3-4, also exerting distinctive regional impacts.
M. N. Raphael (2007) shows that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration.

The Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope (Irving and Simmonds 2015).
The latter makes possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
With this approach, Irving and Simmonds (2015) showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.

Another extensively used approach to characterize the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation (Fogt and Marshall 2020).
SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast (Mo and Paegle 2001).
These patterns influence precipitation anomalies in South America (Mo and Paegle 2001).
Although these patterns are usually derived by applying EOF to temporal anomalies, (M. Raphael 2003) also applied EOF methods specifically to zonal anomalies.
More recently, (Irving and Simmonds 2016) proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.

Patterns resulting from EOF analyses are more flexible than Fourier decomposition derived modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying features such as zonal waves.
A single EOF can also represent a mixture of two or more physical modes.

A third methodology commonly used consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
Examples of this methodology are the SAM Index of (Gong and Wang 1999), the SH wave 3 activity index defined by M. N. Raphael (2004) and the SH zonally asymmetric circulation index from Hobbs and Raphael (2010).
These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
It is commonly easy the computation of these kind of indices, but they could be unable to capture non-stationary patterns.

An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions (Horel 1984).
This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain on water level to study tide-river interactions (Laurel-Castillo and Valle-Levinson 2020) and temperature and precipitation anomalies (e.g. Krokhin and Luxemburg 2007; Gelbrecht, Boers, and Jürgen Kurths 2018).
To our knowledge, complex EOF has not been applied in the spatial domain to capture the phase-varying nature of planetary waves.

Some of the zonally asymmetric patterns of the SH circulation variability described previously, appear to have experienced secular changes.
For instance, M. Raphael (2003) found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
However, little is known yet about variability and trends of these patterns.

The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using complex Empirical Orthogonal Functions, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
In addition, it is proposed to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.

We restrict this paper to the September-October-November (SON) trimester.
We chose this season because it showed the strongest and more clearly interpretable results in a preliminary analysis of the four season.
This is is consistent with previous research that shows that this season experiences a maximum in tropical teleconnections (Cazes-Boezio, Robertson, and Mechoso 2003) and stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).

In Section 2 we describe the methods.
In Section 3.1, we analyse the spatial patterns of each complex EOF.
In Section 3.2 we study the spatial regressions with geopotential height and temperature anomalies.
In Section 3.3, we analyse the relationship between cEOF1 and Ozone, and between cEOF2 with the SAM and the PSA.
In Section 3.4 we study tropical forcings that explain the variability of each cEOF.
Finally, in Section 3.5 we show the relationship between these modes of variability and precipitation anomalies in South America and Oceania.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; Hersbach et al. (2019){]} for the period 1979-2019.
Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 (Bell et al. 2020) to describe long-term trends.
We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK (Adams, Swartztrauber, and Sweet 1999) and we computed horizontal wave activity fluxes following Plumb (1985).
Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 (Huang et al. 2017) and precipitation monthly data from the CPC Merged Analysis of Precipitation (P. Xie and Arkin 1997), with a 2.5\degree~resolution in latitude and longitude.
The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

@@

The amplitude of the zonal waves was obtained through the Fourier transform of the spatial field at each latitude circle.
For the analysis of wave 1, we computed its amplitude and phase by averaging (area-weighted) the data for each variable and each SON between 75°S and 45°S, and then extracting the wave-1 component of the Fourier spectrum.
We chose this latitude band because it is wide enough to capture most of the relevant anomalies of SH mid-latitudes.

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according (Baldwin and Thompson 2009).
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022).
Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated -- in agreement with Mo and Paegle (2001) -- as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height.

Linear trends were computed by Ordinary Least Squares (OLS) and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom (D. Wilks 2011).

@@

\begin{figure}
\centering
\includegraphics{../figures/eof-naive-1.pdf}
\caption{\label{fig:eof-naive}Spatial patterns of the four leading EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).}
\end{figure}

In standard way of applying EOF analysis zonal waves may appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase (Horel 1984).
Figure~\ref{fig:eof-naive} shows the four eading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
A similar EOF structure can be seen in 200 hPa (not shown).
Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

ON the other hand, complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, associated with phase-varying structures (Horel 1984).
This method involves computing EOF on the analytic representation of the original field.
That representation is a complex field in which the real part is the original data and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
The Hilbert transform is usually understood in terms of time-varying signal.
However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.

The result of the cEOF methodology is a set of complex spatial patterns and complex time series.
The real and imaginary part of each spatial pattern represent the two phases of wave-like spatial patterns that are in quadrature.
The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.

@@

\caption{\label{tab:corr-ceof-splitted}Coefficient of determination (\(R^2\)) between the absolute magnitude of complex EOFs computed separately at 200 hPa and 50 hPa (p-values lower than 0.01 in bold).}
\centering
\begin{tabular}[t]{l>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{} & \multicolumn{3}{c}{50 hPa} \\
\cmidrule(l{3pt}r{3pt}){2-4}
200 hPa & cEOF1 & cEOF2 & cEOF3\\
\midrule
cEOF1 & \textbf{0.29} & 0.01 & 0.03\\
cEOF2 & 0.00 & \textbf{0.59} & 0.02\\
cEOF3 & 0.00 & 0.00 & 0.01\\
\bottomrule
\end{tabular}
\end{table}

The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S separately at 50 and 200 hPa.
Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
There is a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).

Both the spatial pattern similarity and the high temporal correlation of cOEFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability.
This motivates the decision of performing complex EOF jointly between levels.
Therefore cEOFs were computed using data from both levels at the same time.
In that sense each cEOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

The phase of principal components is defined up to an additive constant.
For real principal components, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
For complex principal components, it can be any number between 0 and \(2\pi\) (Horel 1984).
Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
This procedure does not create a spurious correlation, it only takes a known relationship and aligns it with a specific phase.
For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
This also nearly minimises the correlation with the imaginary part.
For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.

@@

We performed linear regressions to quantify the association between the cEOFs and other variables (e.g.~geopotential height, temperature, precipitation, and others).
For each cEOF, we computed regression maps by fitting a multiple linear model involving both the real and the imaginary part.
To obtain the linear coefficients of a variable \(X\) with the imaginary and real parts of each cEOF we fit the equation

@@

where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients for imaginary and real parts respectively, \(X_0\) and \(\epsilon\) are the constant and error terms respectively.

We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate (Benjamini and Hochberg 1995; D. S. Wilks 2016) to avoid misleading results from the high number of regressions (Walker 1914; Katz and Brown 1991).

@@

We performed all analysis in this paper using the R programming language (R Core Team 2020), using data.table (Dowle and Srinivasan 2020) and metR (Campitelli 2020) packages. All graphics are made using ggplot2 (Wickham 2009). We downloaded data from reanalysis using the ecmwfr package (Hufkens 2020) and indices of the ENSO with the rsoi package (Albers and Campitelli 2020). The paper was rendered using knitr and rmarkdown (Y. Xie 2015; Allaire et al. 2020).

@@

\begin{figure*}
\includegraphics{../figures/extended-series-1} \caption{Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019). Dark straight line is the linear trend. Black horizontal and vertical line mark the mean value and range of each time series, respectively.}\label{fig:extended-series}
\end{figure*}

@@

The cEOF1 (Fig.~\ref{fig:ceofs-1} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
The cEOF2 (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
This modulated amplitude is especially apparent at 200 hPa.
There is no apparent phase shift but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.

All cEOFs have non zero mean (Fig. \ref{fig:extended-series}) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
There is no significant simultaneous correlation between the time series.
Both cEOFs show year-to-year variability but show no evidence of decadal variability.

A significant upward trend in the real component of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the complex components of cEOF2.
The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
In agreement, M. Raphael (2003) detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.
